---
title: "EDA. Taras Rodzin"
output:
  html_document:
    df_print: paged
---
```{r}
#install.packages("janitor")
#install.packages("naniar")
#install.packages("ggpubr")
#install.packages("plotly")
library(ggplot2)
library(janitor)
library(dplyr)
library(naniar)
library(ggplot2)
library(plotly)
library(ggpubr)
```

## **Read Data**
The first step is to read the data and to pre-process it properly for a further analysis.
```{r}
setwd("D:/Studium/year_two/summer-internships/nyu-ml-course") 
raw_data <- as.data.frame(read.table("D:/Studium/year_two/summer-internships/nyu-ml-course/fars.dat", header=TRUE, skip = 34, sep= ","))
head(raw_data, 5)
```
## **Distributions Analysis**
```{r}
summary(raw_data) 
```
```{r}
temp <- filter(raw_data, (INJURY_SEVERITY != "Unknown") & (INJURY_SEVERITY != "Unknown") & (SEX != "Unknown") & (TAKEN_TO_HOSPITAL != "Unknown"))
```

### **Get column names**
```{r}
col_labels <- names(raw_data)
features <- col_labels[col_labels != "INJURY_SEVERITY"]
features
```


### **Get numerical & categorical column names**
```{r}
if_numerical <- unlist(lapply(raw_data, is.numeric))
numerical_c <- c(col_labels[if_numerical == TRUE])
categorical_c <- c(col_labels[if_numerical == FALSE])
```

### **Numerical columns**
```{r}
numerical_c
```

### **Categorical columns**
```{r}
categorical_c
```

### **Check If Nan values are present in any column.**
```{r}
cnames <- colnames(raw_data)
for(i in 1:ncol(raw_data)) {
  cat(cnames[i], ": ", sum(is.na(raw_data[ , i])))
  cat("\n")
}
```

### **Check If "Unknown" values are present in any column.**
```{r}
unknowns <- c()
col_labels <- c()

for(i in 1:ncol(raw_data)) {
  feat <- cnames[i]
  n_unknowns <- sum(grepl("Unknown", raw_data[ , i], fixed = TRUE) == TRUE)
  
  cat(feat, ": ", n_unknowns)
  cat("\n")
  
  unknowns <- append(unknowns, n_unknowns)
  col_labels <- append(col_labels, feat)
}
```

### **Plot the distribution of "Unknown" values.**
```{r}
fig <- plot_ly(y=unknowns, x=col_labels, type = 'bar') %>% 
  layout(xaxis = list(categoryorder = "total descending"),
         title = "'Unknown' values distribution among columns")
fig
```

### **Check Class Distribution of Target Variable**
```{r}
fig <- plot_ly(raw_data, labels = ~INJURY_SEVERITY, type = 'pie')
fig <- fig %>% layout(title = 'INJURY_SEVERITY classes distribution',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

fig
```
## **Plot Distribution of Numerical Columns**
### **Plot the distribution of Age feature**
```{r}
dens <- density(raw_data$AGE)


fig <- plot_ly(
  data = raw_data,
  x = ~AGE,
  type = "histogram",
  name = "Age"
) %>% 
  add_lines(x = dens$x, y = dens$y, yaxis = "y2", name = "Density") %>% 
  layout(yaxis2 = list(overlaying = "y", #Adds the dual y-axis
                       side = "right", #Adds the density axis on the right side
                       rangemode = "tozero")) %>%
  layout(title = 'AGE column distribution')
fig
```

### **Plot the distribution of ALCOHOL_TEST_RESULT feature**
```{r}
dens <- density(raw_data$ALCOHOL_TEST_RESULT)


fig <- plot_ly(
  data = raw_data,
  x = ~ALCOHOL_TEST_RESULT,
  type = "histogram",
  name = "Age"
) %>% 
  add_lines(x = dens$x, y = dens$y, yaxis = "y2", name = "Density") %>% 
  layout(yaxis2 = list(overlaying = "y", #Adds the dual y-axis
                       side = "right", #Adds the density axis on the right side
                       rangemode = "tozero")) %>%
  layout(title = 'ALCOHOL_TEST_RESULT column distribution')
fig
```

### **Plot the distribution of DRUG_TEST_RESULTS_.1_of_3. feature**
```{r}
dens <- density(raw_data$DRUG_TEST_RESULTS_.1_of_3.)


fig <- plot_ly(
  data = raw_data,
  x = ~DRUG_TEST_RESULTS_.1_of_3.,
  type = "histogram",
  name = "DRUG_TEST_RESULTS_.1_of_3."
) %>% 
  add_lines(x = dens$x, y = dens$y, yaxis = "y2", name = "Density") %>% 
  layout(yaxis2 = list(overlaying = "y", #Adds the dual y-axis
                       side = "right", #Adds the density axis on the right side
                       rangemode = "tozero")) %>%
  layout(title = 'DRUG_TEST_RESULTS_.1_of_3. column distribution')
fig
```
### **Plot the distribution of DRUG_TEST_RESULTS_.2_of_3. feature**
```{r}
dens <- density(raw_data$DRUG_TEST_RESULTS_.2_of_3.)


fig <- plot_ly(
  data = raw_data,
  x = ~DRUG_TEST_RESULTS_.2_of_3.,
  type = "histogram",
  name = "DRUG_TEST_RESULTS_.2_of_3."
) %>% 
  add_lines(x = dens$x, y = dens$y, yaxis = "y2", name = "Density") %>% 
  layout(yaxis2 = list(overlaying = "y", #Adds the dual y-axis
                       side = "right", #Adds the density axis on the right side
                       rangemode = "tozero")) %>%
  layout(title = 'DRUG_TEST_RESULTS_.2_of_3. column distribution')
fig
```
### **Plot the distribution of DRUG_TEST_RESULTS_.3_of_3. feature**
```{r}
dens <- density(raw_data$DRUG_TEST_RESULTS_.3_of_3.)


fig <- plot_ly(
  data = raw_data,
  x = ~DRUG_TEST_RESULTS_.3_of_3.,
  type = "histogram",
  name = "DRUG_TEST_RESULTS_.3_of_3."
) %>% 
  add_lines(x = dens$x, y = dens$y, yaxis = "y2", name = "Density") %>% 
  layout(yaxis2 = list(overlaying = "y", #Adds the dual y-axis
                       side = "right", #Adds the density axis on the right side
                       rangemode = "tozero")) %>%
  layout(title = 'DRUG_TEST_RESULTS_.3_of_3. column distribution')
fig
```

### **Numerical Columns Distribution with respect to the Target Variable**
```{r}
fig <- plot_ly(temp, x = ~INJURY_SEVERITY, y = ~ALCOHOL_TEST_RESULT, color = ~INJURY_SEVERITY, type = "box") %>%
       layout(title = 'ALCOHOL_TEST_RESULT column distribution regarding INJURY_SEVERITY')
fig
```
```{r}
fig <- plot_ly(temp, x = ~INJURY_SEVERITY, y = ~DRUG_TEST_RESULTS_.1_of_3., color = ~INJURY_SEVERITY, type = "box") %>%
       layout(title = 'DRUG_TEST_RESULTS_.1_of_3. column distribution regarding INJURY_SEVERITY')
fig
```

```{r}
fig <- plot_ly(temp, x = ~INJURY_SEVERITY, y = ~DRUG_TEST_RESULTS_.2_of_3., color = ~INJURY_SEVERITY, type = "box") %>%
       layout(title = 'DRUG_TEST_RESULTS_.2_of_3. column distribution regarding INJURY_SEVERITY')
fig
```

```{r}
fig <- plot_ly(temp, x = ~INJURY_SEVERITY, y = ~DRUG_TEST_RESULTS_.3_of_3., color = ~INJURY_SEVERITY, type = "box") %>%
       layout(title = 'DRUG_TEST_RESULTS_.3_of_3. column distribution regarding INJURY_SEVERITY')
fig
```

## **Plot Distribution of Categorical Columns**
```{r}

for (i in 1:length(col_labels)) {
  
  if (col_labels[i] %in% categorical_c) {
    fig <- plot_ly(raw_data, labels = raw_data[,i], type = 'pie')
    fig <- fig %>% layout(title = sprintf("%s Distribution", col_labels[i]),
             xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
             yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
    
    print(fig)
  }
}

```

### **Plot SEX Column Distribution with Respect to Target Variable**
```{r}
#create table of counts
counts2 <- table(temp$INJURY_SEVERITY, temp$SEX)

#create mosaic plot
subs_pal = colorspace::qualitative_hcl(2)
mosaicplot(counts2, xlab='INJURY_SEVERITY', ylab='SEX',
           main='INJURY_SEVERITY & SEX', col=subs_pal, off = 10, las = 1)

```

### **Plot TAKEN_TO_HOSPITAL Column Distribution with Respect to Target Variable**
```{r}
#create table of counts
counts1 <- table(temp$INJURY_SEVERITY, temp$TAKEN_TO_HOSPITAL)

#create mosaic plot
subs_pal = colorspace::qualitative_hcl(2)
mosaicplot(counts1, xlab='INJURY_SEVERITY', ylab='TAKEN_TO_HOSPITAL',
           main='INJURY_SEVERITY & TAKEN_TO_HOSPITAL', col=subs_pal, off = 10, las = 1)
```

### **Categorical Columns Distribution with respect to the Target Variable**
```{r}
#create table of counts
for (i in 1:length(col_labels)) {
  
  if (col_labels[i] %in% categorical_c) {
    print("")
    print(col_labels[i])
    print(table(temp$INJURY_SEVERITY, temp[, i]))
    print("")
  }
}
```

## **Plot Correlations**
```{r}
for (i in 1:length(names(raw_data))) {
  x = as.numeric(factor(raw_data[,i]))
  raw_data[,i] = x
}
data_corr <- cor(raw_data)
```

### **Plot Correlation Matrix**
```{r}
fig <- plot_ly(x = names(data.frame(data_corr)), 
               y = names(data.frame(data_corr)), 
               z = data_corr, type = "heatmap") %>% 
  layout(title = "Correlation Within the Dataset", autosize = F, width = 1000, height = 1000)

fig
```

### **Features Ordered by Corr value with Target Variable**
```{r}
target_corr <- c(data.frame(data_corr)$INJURY_SEVERITY)
target_cols <- c(names(data.frame(data_corr)))

top_ten_cols <- target_cols[order(abs(target_corr), decreasing=TRUE)]
top_ten_corr <- target_corr[order(abs(target_corr), decreasing=TRUE)]

# Sort the attributes by correlation value and exclude the 1st one (target)
print("Top 10 attributes correlated (absolute value) with INJURY_SEVERITY:")
for (i in 2:length(top_ten_cols)) {
  print(paste(i-1,".", top_ten_cols[i], "  -->  ", top_ten_corr[i]))
}
```